---
title: Choose your favorite color
author: Package Build
date: '2024-09-20'
slug: choose-your-favorite-color
categories: []
tags:
  - R
  - ggplot2
  - interactivity
comments: no
images: ~
---

The function below provides quick access to R color name information in an R Studio session. Running the function provides a grid of tiled colors, where you may hover over the tile to get its R color name, hex code, and rgb value:

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(ggiraph)
library(grDevices)

get_hsv_values <- function(color_name) {
  rgb_values <- col2rgb(color_name) / 255
  hsv_values <- rgb2hsv(rgb_values[1], rgb_values[2], rgb_values[3])
  tibble(H = hsv_values['h', ], S = hsv_values['s', ], V = hsv_values['v', ])
}

plot_colors <- function() {
  colors_df <- tibble(
    color_name = colors(),
    hex_code = rgb(
      col2rgb(colors())[1, ],
      col2rgb(colors())[2, ],
      col2rgb(colors())[3, ],
      maxColorValue = 255
    )
  ) |>
    distinct(hex_code, .keep_all = TRUE) |>
    mutate(hsv_value = map(color_name, get_hsv_values)) |>
    unnest(hsv_value) |>
    arrange(H, S, V)
  
  n_colors <- nrow(colors_df)
  grid_size <- ceiling(sqrt(n_colors))
  colors_df <- colors_df |>
    mutate(
      x = rep(1:grid_size, length.out = n_colors),
      y = rep(1:grid_size, each = grid_size)[1:n_colors]
    )
  
  p <- ggplot(colors_df, aes(x, y)) +
    geom_tile_interactive(
      aes(
        fill = color_name,
        tooltip = paste0(
          '<table><tr><td><b>Color:</b></td><td>',
          color_name,
          '</td></tr>',
          '<tr><td><b>Hex:</b></td><td>',
          hex_code,
          '</td></tr>',
          '<tr><td><b>RGB:</b></td><td>',
          paste(
            col2rgb(color_name)[1, ],
            col2rgb(color_name)[2, ],
            col2rgb(color_name)[3, ],
            sep = ', '
          ),
          '</td></tr></table>'
        ),
        data_id = color_name
      ),
      color = 'black',
      width = 0.95,
      height = 0.95
    ) +
    annotate(
      'text',
      x = 20,
      y = 22,
      label = ' > colors()',
      hjust = 0,
      size = 11 / .pt
    ) +
    scale_fill_manual(values = setNames(colors_df$hex_code, colors_df$color_name)) +
    theme_void() +
    theme(legend.position = 'none') +
    labs(title = 'Hover over a tile to reveal its name, hex code, and rgb values.')
  
  girafe(
    ggobj = p,
    options = list(
      opts_selection(type = 'single'),
      opts_hover(css = 'fill-opacity:0.8;'),
      opts_sizing(rescale = TRUE)
    )
  )
}

plot_colors()

```

To have function ready when you open R Studio, place the code into a file, say, `plot_colors.r` and follow these steps:

## Instructions: 

### Save the file: 

Make sure the `plot_colors.r` file (which contains the code) is saved in
a location where it can be easily accessed.

### Locate or create the `.Rprofile` file: 

The `.Rprofile` file is an initialization script that R reads every time
it starts. This file can either be in the project's root directory (if
you want it to load only for that project) or in your home directory (if
you want it to load globally every time RStudio starts, regardless of
the project).

To create or locate the global `.Rprofile`, use the following command in
R: `file.edit("~/.Rprofile")` This will either open the existing
`.Rprofile` or create a new one in your home directory.

### Modify the .Rprofile file: 

In the `.Rprofile` file, add the following line of code to source the
`plot_colors.r` file when RStudio starts:
`source("/path/to/your/plot_colors.r")`

Make sure to replace "`/path/to/your/plot_colors.r`" with the actual
file path where `plot_colors.r` is located.

If you are working in a specific RStudio project and want it to load
only for that project, create a `.Rprofile` file in the root directory
of the project and add the same line.

### Restart RStudio: 

Once you’ve saved the changes to the .Rprofile file, restart RStudio.
The code from `plot_colors.r` should automatically load into your R
environment each time you start RStudio.

### Verifying the Setup: 

After RStudio restarts, you can use the function `plot_colors()` With
these steps, the code in `plot_colors.r` will be sourced into the R
environment every time RStudio starts.

## Code:

```{r}
#| eval: false
library(tidyverse)
library(ggiraph)
library(grDevices)

get_hsv_values <- function(color_name) {
  rgb_values <- col2rgb(color_name) / 255
  hsv_values <- rgb2hsv(rgb_values[1], 
                        rgb_values[2], 
                        rgb_values[3])
  tibble(
    H = hsv_values['h', ], 
    S = hsv_values['s', ], 
    V = hsv_values['v', ])
}

plot_colors <- function() {
  colors_df <- tibble(
    color_name = colors(),
    hex_code = rgb(
      col2rgb(colors())[1, ],
      col2rgb(colors())[2, ],
      col2rgb(colors())[3, ],
      maxColorValue = 255
    )
  ) |>
    distinct(hex_code, .keep_all = TRUE) |>
    mutate(
      hsv_value = map(color_name, get_hsv_values)
    ) |>
    unnest(hsv_value) |>
    arrange(H, S, V)
  
  n_colors <- nrow(colors_df)
  grid_size <- ceiling(sqrt(n_colors))
  colors_df <- colors_df |>
    mutate(
      x = rep(1:grid_size, length.out = n_colors),
      y = rep(1:grid_size, each = grid_size)[1:n_colors]
    )
  
  p <- ggplot(colors_df, aes(x, y)) +
    geom_tile_interactive(
      aes(
        fill = color_name,
        tooltip = paste0(
          '<table><tr><td><b>Color:</b></td><td>',
          color_name,
          '</td></tr>',
          '<tr><td><b>Hex:</b></td><td>',
          hex_code,
          '</td></tr>',
          '<tr><td><b>RGB:</b></td><td>',
          paste(
            col2rgb(color_name)[1, ],
            col2rgb(color_name)[2, ],
            col2rgb(color_name)[3, ],
            sep = ', '
          ),
          '</td></tr></table>'
        ),
        data_id = color_name
      ),
      color = 'black',
      width = 0.95,
      height = 0.95
    ) +
    annotate('text', x = 20, y = 22, label = ' > colors()',
      hjust = 0, size = 11 / .pt
    ) +
    scale_fill_manual(
      values = setNames(colors_df$hex_code, colors_df$color_name)
    ) +
    theme_void() +
    theme(legend.position = 'none') +
    labs(
      title = str_c('Hover over a tile to reveal',
                    'its name, hex code, and rgb values.')
      )
  
  girafe(
    ggobj = p,
    options = list(
      opts_selection(type = 'single'),
      opts_hover(css = 'fill-opacity:0.8;'),
      opts_sizing(rescale = TRUE)
    )
  )
}

plot_colors()

```

