---
title: Tables in tooltips
author: Scott Spencer
date: '2020-06-07'
slug: tables-in-tooltips
categories: []
tags:
  - Communiction
  - R
  - ggplot2
  - grammar of graphics
  - interactivity
comments: no
images: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  cache = FALSE, 
  eval = TRUE)
```

With interactive communications, we frequently encourage the graphic to provide some form of *overview*, and use interactions for adding *details on demand*. For these details, tables provide a flexible way to organize the information. And we can link the view of such tables to user actions like a *pointer hover event* where the table of details is in a *tooltip*. In this tutorial, we construct an `html` table for our tooltip so that we can organize lots of details on demand.

Such tooltips are easy to customize within the `R` ecosystem, especially when founded upon or extending the *grammar of graphics*. Here, our code relies on functions in `ggplot2`, `ggiraph`, and `flextable`. Let's load these first.

```{r}
library(tidyverse)
library(flextable)
library(ggiraph)
```

For our example data, we construct a periodic table of elements where the hover events adds details about each element. Information on the known elements are available in another package, `rbokeh`. Next, we load this data, 

```{r}
data(list = 'elements', package = 'rbokeh')
```

and transform it a bit. First, we define a color for each element type:

```{r}
element_type <- c(
  "alkali metal", "alkaline earth metal", "halogen", "metal", 
  "metalloid", "noble gas", "nonmetal", "transition metal")

element_color <- c(
  "#a6cee3", "#1f78b4", "#fdbf6f", "#b2df8a", 
  "#33a02c", "#bbbb88", "#baa2a6", "#e08e79")
```

Next, we add this information to our data:

```{r}
elements <- 
  elements %>%
  filter( !is.na(group) ) %>%
  mutate(
    across(.cols = group:atomic.mass, as.character),
    atomic.mass = if_else(symbol == "Uus", " ", atomic.mass),
    color = element_color[match(metal, element_type)],
    type = metal
  )
```

Our data are structured as follows:

```{r}
cbind(names(elements), sapply(elements, class)) %>% 
data.frame %>% 
flextable(col_keys = c('X1', 'X2')) %>%
set_header_labels( X1 = 'Variable', X2 = 'Type' ) %>%
color(color = '#777777', part = 'all')
```

Next, we create a table for each element. To do this, we create a function to build an `html` table from whatever data variables we pass it, returning the html code as a string. then, for each element (each row in our dataframe), we save the table as a new variable in the dataframe (we'll call the variable `tooltip`) that holds a string of the `html` defining the table. Here's the function:

```{r}
tooltip_table <- function(...) {
  
  data.frame(...) %>% 
  mutate( across(.cols = everything(), as.character) ) %>%
  pivot_longer(cols = everything(), names_to = 'P', values_to = 'V') %>%
  flextable(col_keys = c('P', 'V')) %>%
  set_header_labels( P = '', V = '' ) %>%
  fontsize(size = 8, part = 'all') %>%
  line_spacing(space = 5, unit = 'mm', part = 'all') %>%
  set_table_properties(layout = 'autofit') %>%
  theme_alafoli() %>%
  htmltools_value(ft.shadow = FALSE) %>%
  as.character()
  
}
```

Notice that all we do is convert whatever variables the function receives (`...`) into a data.frame (`data.frame(...)`), coerce them all to character strings (`mutate()`), pivot them to long format (`pivot_longer()`) so that we can list each variable alongside its value, and then create the table, formatted however we like using the syntax described in the [book for FlexTable](https://ardata-fr.github.io/flextable-book/). Of note, while we only list variables with their values, we can extend our table to include graphics, too. See the documentation.

Next, as mentioned, we save the html code for each observation as a string in the variable `tooltip`.

```{r}
elements <- 
  elements %>%
  rowwise() %>%
  mutate(tooltip = tooltip_table( 
    name, type, atomic.mass, density, electronic.configuration, 
    boiling.point, melting.point, bonding.type) 
  ) %>%
  ungroup()
```

Finally, now that we have the html code to construct a table for each unique data observation, we can link it to the `tooltip` parameter in any interactive geometry from `ggiraph` that maps data (the variable we called `tooltip`) to the function parameter by the same name. Here is the grammar of graphics specification to construct the interactive table:

```{r}
p <- 
  elements %>%
  
  ggplot(
    mapping = aes(group, period)
  ) +
  
  theme_void() +
  
  coord_equal() +
  
  scale_x_discrete(limits = factor(1:18) ) +
  scale_y_discrete(limits = factor(7:1) ) +
  scale_fill_identity() +
  
  # base color of each element square
  geom_tile(
    mapping = aes(
      fill = color,
    ),
    alpha = 0.6,
    width = 1,
    height = 1,
    color = '#ffffff',
    lwd = 0.3
  ) +
  
  # add symbol to each square
  geom_text_interactive(
    mapping = aes(
      label = symbol,
      data_id = atomic.number
    ),
    size = 7/.pt,
    hjust = 0,
    vjust = 0.5,
    nudge_x = -0.4,
    nudge_y = 0.04,
    fontface = 'bold'
  ) +
  
  # add atomic number to each square
  geom_text_interactive(
    mapping = aes(
      label = atomic.number,
      data_id = atomic.number
    ),
    size = 4/.pt,
    hjust = 0,
    vjust = 0.5,
    nudge_x = -0.4,
    nudge_y = 0.35
  ) +
  
  # add name to each square
  geom_text_interactive(
    mapping = aes(
      label = name,
      data_id = atomic.number
    ),
    size = 3/.pt,
    hjust = 0,
    vjust = 0.5,
    nudge_x = -0.4,
    nudge_y = -0.2
  ) +
  
  # add atomic mass to each square
  geom_text_interactive(
    mapping = aes(
      label = atomic.mass,
      data_id = atomic.number
    ),
    size = 3/.pt,
    hjust = 0,
    vjust = 0.5,
    nudge_x = -0.4,
    nudge_y = -0.35
  ) +
  
  # for mouseover, set alpha very low
  geom_tile_interactive(
    mapping = aes(
      fill = color,
      data_id = atomic.number,
      tooltip = tooltip
    ),
    alpha = 0.01,
    width = 1,
    height = 1,
    color = '#ffffff',
    lwd = 0.3
  )
```

Of note, we include the last `geom_tile_interactive` because we construct the grammar of graphics in layers, and the pointer hover will interact with whatever the top layer is. By including a top layer as an almost fully transparent tile, the pointer will interact with the entire square of information.

Finally, we bind the interactive graphic to interactive components and specify css for the interactivity.

```{r}
girafe(
  ggobj = p,
  options = list(
    opts_hover(
      css = 'fill-width:3;'),
    opts_hover_inv(
      css = 'fill-opacity:0.2;'),
    opts_tooltip(
      css = str_c(
        'background-color:white;',
        'color:black;',
        'font-size:8pt;',
        'padding:10px;'
      ),
      offx = 10,
      offy = 10
    )
  )
)
```

As mentioned, we can extend this idea in limitless ways. I hope this gives you ideas for your own work.

Stay curious.
