---
title: "Exploratory modeling on how win uncertainty affects baseball game attendance"
date: 2020-01-03T18:01:28-04:00
draft: false
comments: false
tags:
  - R
  - Stan
  - Baseball
  - Uncertainty
  - Betting
  - Attendance
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE)
```

We dread an empty ticket booth. The Dodgers' average game attendance, for example, can be below capacity — they have seats to fill, despite having winning seasons.

To draw more fans, consider the words of famed economist Simon Rotternberg, who said “uncertainty of outcome is necessary for consumers to pay admission to ball games." While winning is important, it’s the not knowing that creates excitement. Have the Dodgers maximized winning with uncertainty of outcome? Let's see.

To start exploring the question, we model attendance as a function of the uncertainty of outcome, attempting to account for other factors like day of the week, time of day, and their cumulative fraction of wins. It seems reasonable to believe that some uncertainty helps attract people to the game. It also seems reasonable to believe that the function is non-linear: _a change in probability of a win from 0 percent to 1 percent may well attract fewer fans than if from 49 percent to 50 percent._ Thus, let's try modeling the marginal effect of wins as quadratic. Our overall model, then, looks like:

$$\textrm{Normal}(\theta, \sigma)$$ 

for game $i$, where 

$$\theta_i \sim \alpha_{1[i]} \cdot \textrm{day}_i + \alpha_{2[i]} \cdot \textrm{time}_i + \beta_{1[i]} \cdot \frac{\sum{\textrm{wins}_i}}{\sum{\textrm{games}_i}} + \beta_{2[i]} \cdot p(\textrm{win}_i) + \beta_{3[i]} \cdot p(\textrm{win}_i)^2$$
Once we get our posterior estimates, we can set the partial derivative of estimates of win uncertainty ($\beta_2$ and $\beta_3$) to find a maximum:

$$\textrm{Maximum} = \frac{-\beta_2}{2 \cdot \beta_3 }$$
We'll run the analysis in `Stan`.

For our analysis, let's use betting market odds as a proxy for fans' estimation their teams chances of winning. The betting company Pinnacle has made these data available for the 2016 season in their R package `pinnacle.data::MLB2016`. As with Pinnacle's generosity, [Retrosheets has made data available for per-game attendance along with game outcomes](https://www.retrosheet.org/gamelogs/index.html).

```{r echo=FALSE}
library(dplyr)          # organize data
library(odds.converter) # calculate betting percentages
library(rstanarm)       # fit model
library(ggplot2)        # make graphic
library(ggthemes)       # tufte theme settings on graphic
library(latex2exp)      # format latex equations on graphic
```

```{r echo=FALSE}
betting <- pinnacle.data::MLB2016 %>% # betting data available in package pinnacle.data
  tidyr::unnest(cols = c(Lines)) %>%  # convert nested list to data.frame
  group_by(GameID) %>%                # there are multiple betting data per game, so
  arrange(EventDateTimeUTC) %>%       # arrange in time bets placed
  slice(n()) %>%                      # and only keep last betting odds in each game
  ungroup()                           # generally good to ungroup what you've grouped
```

```{r echo=FALSE}
GL2016 <- read.csv("data/GL2016.csv", header = FALSE, stringsAsFactors = FALSE)
gamelog_header <- read.csv("data/gamelog_header.csv")
names(GL2016) <- names(gamelog_header)
GL2016 <- GL2016 %>% 
  mutate(GameID = paste0(HomeTeam, Date, DoubleHeader))
```

```{r echo=FALSE}
df <- inner_join(betting, GL2016, 
                 by = c("GameID" = "GameID"))
df <- df %>% arrange(GameID)
```

```{r echo=FALSE}
dodgers2016 <- df %>% 
  filter(HomeTeam.y == "LAN") %>% # only consider Dodgers home games
  select(Date, Attendence, MoneyUS1, MoneyUS2,  # filter only columns plan to use
         DayOfWeek, DayNight, FinalScoreHome, FinalScoreAway) %>% 
  mutate(is_win = FinalScoreHome > FinalScoreAway) %>% # calculate flag whether Dodgers won
  mutate(winpct = cumsum(is_win) / row_number()) %>% # calculate win percent 
  mutate(winpct = lag(winpct)) %>% # get win percent of to prior game
  filter(complete.cases(winpct))
```

```{r echo=FALSE}
dodgers2016$Pr_win <- 1 - with(dodgers2016, {
  Pr <- t(apply(cbind(MoneyUS1, MoneyUS2), 1, FUN = odds.us2prob))
  Pr <- Pr / rowSums(Pr)
  Pr[,1]
})
```

```{r echo=FALSE}
dodgers2016$Pr_win2 <- dodgers2016$Pr_win * dodgers2016$Pr_win
```

In code, our basic model is as follows:

```{r}
fit <- stan_glm(Attendence ~ 1 + DayOfWeek + DayNight + winpct + Pr_win + Pr_win2, 
                data = dodgers2016,
                cores = 4, chains = 4)
```

```{r echo=FALSE}
post <- data.frame(fit)
```

```{r echo=FALSE}
uncertainty <- with(post, (-Pr_win) / (Pr_win2 * 2))
```

Our resulting estimate for the distribution of the marginal perception of outcome uncertainty that maximizes butts in seats was a median of `r round(quantile(uncertainty, probs = 0.5), 2)` with the distribution:

```{r echo=FALSE}
scaleFUN <- function(x) sprintf("%.2f", x)

eqn <- paste0("$Normal(\\theta, \\sigma)$" , ", ", 
              "$\\theta \\sim \\alpha_1 \\cdot day + \\alpha_2 \\cdot time + ",
              "\\beta_1 \\cdot \\frac{\\sum wins}{\\sum games} + ",
              "\\beta_2 \\cdot p(win) + \\beta_3 \\cdot p(win)^2$")


eqn2 <- paste0("$Maximum = \\frac{-\\beta_2}{2 \\cdot \\beta_3}$")

# plot graphic with annotations for specific audience
ggplot(x = uncertainty) + 

  # data-ink
  geom_density(aes(uncertainty), 
               fill = "#bbbbbb", 
               color = "#aaaaaa") + 

  # non-data ink

  # change default theme attributes
  theme_tufte(base_family = "sans") +
  
  # cut off impossible values (probability < 0 or > 1), set x tick labels at specific amounts relevant
  # to understanding, and format them only to two digits
  scale_x_continuous(limits = c(0, 1), 
                     breaks = c(0, 
                                .5, 
                                quantile(uncertainty, 
                                         probs = c(.1, .5, .9)), 
                                1), 
                     labels = scaleFUN) +
  
  # just to remove padding above, below data
  scale_y_continuous(expand = c(0, 0)) +
  
  # various labels and annotations
  # lighten lower quantile of data
  annotate("rect", 
           xmin = 0, 
           ymin = 0, 
           xmax = quantile(uncertainty, probs = .1), 
           ymax = 5, 
           fill = "#ffffff", 
           alpha = .5) +
  
  # lighten upper quantile of data
  annotate("rect", 
           xmin = quantile(uncertainty, probs = .9), 
           ymin = 0, 
           xmax = 1, 
           ymax = 5, 
           fill = "#ffffff", 
           alpha = .5) +
  
  # Draw vertical white lines on quantiles
  geom_vline(xintercept = quantile(uncertainty, 
                                   probs = c(.1, .5, .9)), 
             color = "#ffffff") +
  
  # draw vertical blue line on 50% or coin flip and label it as such
  geom_vline(xintercept = .5, 
             color = "#407AAA", 
             linetype = "dashed") +
  
  annotate("text", 
           x = .496, 
           y = 4.2, 
           label = "coin", 
           size = 5, 
           color = "#000000", 
           hjust = 1, 
           fontface = "bold") +
  
  annotate("text", 
           x = .496, 
           y = 4, 
           label = "loss",
           size = 5, 
           color = "#407AAA", 
           hjust = 1, 
           fontface = "bold") +
  
  annotate("text", 
           x = .504, 
           y = 4.2, 
           label = "flip", 
           size = 5, 
           color = "#000000", 
           hjust = 0, 
           fontface = "bold") +
  
  annotate("text", 
           x = .504, 
           y = 4, 
           label = "win", 
           size = 5, 
           color = "#407AAA", 
           hjust = 0, 
           fontface = "bold") +
  
  # annotate median of probability
  annotate("text", 
           x = median(uncertainty) + .01, 
           y = .5, 
           label = "Probability of\nteam win for\nmaximum attendance", 
           hjust = 0, 
           size = 2.5, 
           fontface = "bold",
           lineheight = .8) +
  
  # # annotate equations
  # annotate("text", 
  #          x = 0, y = 1, 
  #          label = TeX(eqn), 
  #          size = 2.8, 
  #          color = "#aaaaaa", 
  #          hjust = 0) +
  # 
  # annotate("text", 
  #          x = 0, 
  #          y = 0.5, 
  #          label=TeX(eqn2), 
  #          size = 2.8, 
  #          color = "#aaaaaa", 
  #          hjust = 0) +
  
  # additional overall theme (non-data-ink) changes
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 12, face = "bold", color = "#666666"),
        axis.title.y = element_blank(), 
        plot.title = element_text(size = 14, color = "#888888", face = "bold"),
        plot.subtitle = element_text(size = 11, color = "#999999", face = "italic"),
        plot.caption = element_text(size = 8, color = "#888888")) +
  
  # labeling for graphic
  labs(x = "", 
       title = paste0("More fans generally pay admission to our games when\n",
                      "the chance of winning was near a median of ", 
                      signif(median(uncertainty), 2), "."),
       subtitle = "Fans want favorable odds without predicting the outcome.",
       caption = "Sources: Pinnacle betting data, 2016; Retrosheet Gamelogs, 2016")
```

