---
output: 
  bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

<style>

.main-container {
  min-width: 950px;
  max-width: 950px;

}

.gridlayout { 
  display: grid;
  position: relative;
  margin: 10px;
  gap: 5px;
  grid-template-columns: 
    repeat(8, 1fr);
  grid-template-rows: 
    repeat(8, 140px);
}

.title {
  background: lightgray;
  grid-column: 1 / 9;
  grid-row: 1 / 2;
  font-size: 14pt;
}

.large-left {
  background: lightblue;
  grid-column: 1 / 5;
  grid-row: 2 / 5;
}

.medium-middle {
  background: lightblue;
  grid-column: 5 / 7;
  grid-row: 2 / 5;
}

.medium-right {
  background: lightblue;
  grid-column: 7 / 9;
  grid-row: 2 / 5;
}

.bottom-left {
  background: pink;
  grid-column: 1 / 5;
  grid-row: 5 / 8;
}

.bottom-middle {
  background: pink;
  grid-column: 5 / 7;
  grid-row: 5 / 8;
}

.bottom-right {
  background: pink;
  grid-column: 7 / 9;
  grid-row: 5 / 8;
}

</style>

<div class="gridlayout">

<div class="title">This is sample content placed in the area defined by our title class.</div>

<div class="large-left">This is sample content placed in the area defined by our large-left class.
```{r}
plot(pressure)
```

</div>

<div class="medium-middle">This is sample content placed in the area defined by our medium-middle class.
```{r}
plot(pressure)
```
</div>

<div class="medium-right">This is sample content placed in the area defined by our medium-right class.
```{r}
plot(pressure)
```
</div>

<div class="bottom-left">This is sample content placed in the area defined by our bottom-full class.
```{r}
plot(pressure)
```
</div>

<div class="bottom-middle">This is sample content placed in the area defined by our bottom-middle class.
```{r}
plot(pressure)
```
</div>

<div class="bottom-right">This is sample content placed in the area defined by our bottom-right class.
```{r}
plot(pressure)
```
</div>


</div>
